---
output: github_document
---

# densityLP

**Local Polynomial Density Estimation on Complicated Domains**

`densityLP` implements the nonparametric local polynomial density estimator of
Bertin, Klutchnikoff and Ouimet (*Bernoulli*, forthcoming) for multivariate
densities on known domains of arbitrary dimension, including domains with sharp
concavities, holes, and polynomial sectors. The C++ backend
(RcppArmadillo) provides a ×10 speedup over a pure-R implementation.

## Installation

```r
# Development version
remotes::install_github("nklutchnikoff/densityLP")
```

## Quick start

### Local linear estimation on the unit square

```{r square-example}
library(densityLP)

set.seed(1L)
n <- 500L
X <- matrix(runif(n * 2L), n, 2L)   # uniform on [0,1]^2

is_square <- function(x, y) x >= 0 & x <= 1 & y >= 0 & y <= 1
dom       <- domain_from_indicator(is_square)

grid  <- seq(0.1, 0.9, length.out = 20L)
t_grid <- as.matrix(expand.grid(x = grid, y = grid))

fit <- density_lp(X, t_grid = t_grid, h = 0.2, m = 1L,
                  domain = dom, N_quad = 500L)
plot(fit, main = "Local linear estimate — unit square")
```

### Spatial point pattern on an L-shaped domain

`density_lp_ppp()` accepts any `ppp` object from
[spatstat](https://spatstat.org). It derives the evaluation grid and the
integration domain automatically from `Window(pp)` and returns a
`spatstat.geom::im` pixel image.

```{r L-example}
library(spatstat.geom)
library(spatstat.random)

# L-shaped window and inhomogeneous Poisson process
L_win <- owin(poly = list(x = c(0, 2, 2, 1, 1, 0),
                          y = c(0, 0, 1, 1, 2, 2)))

lambda_fun <- function(x, y) {
  400 * exp(-4 * ((x - 1.5)^2 + (y - 0.4)^2)) +
    150 * exp(-4 * ((x - 0.3)^2 + (y - 1.5)^2))
}

set.seed(2L)
pp  <- rpoispp(lambda_fun, win = L_win)
fit <- density_lp_ppp(pp, h = 0.3, m = 1L, N_quad = 500L)

plot(fit, main = "Local linear estimate — L-shaped domain")
plot(pp, add = TRUE, pch = ".", col = "#00000050")
```

Pixels outside the window are automatically set to `NA`.

## Vignettes

| Vignette | Description |
|---|---|
| `convergence` | MSE convergence study on the cube and a polynomial sector |
| `benchmark` | R vs C++ timing: `gram_matrix` and full pipeline |
| `spatstat` | `density_lp_ppp` on synthetic spatial point patterns |

## Reference

Bertin, K., Klutchnikoff, N. and Ouimet, F. (forthcoming).
Local polynomial estimation of the density on complicated domains.
*Bernoulli*.
