---
title: "R vs C++ benchmark: gram_matrix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{R vs C++ benchmark: gram_matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(densityLP)
build_alphas <- densityLP:::build_alphas
gram_matrix_cpp <- densityLP:::gram_matrix_cpp
```

We compare the pure-R implementation of `gram_matrix` (based on
`matrixStats::colProds` + `tcrossprod`) against the C++ backend
(`gram_matrix_cpp`, RcppArmadillo) across varying numbers of quadrature
points $N$ and polynomial degrees $m$.

The two implementations are mathematically identical; the benchmark measures
only wall-clock time.

---

## Setup

```{r params}
library(bench)

d <- 2L # dimension
h <- 0.3 # bandwidth
Ns <- c(100L, 500L, 1000L, 5000L, 10000L)
ms <- c(0L, 1L, 2L, 3L)
```

```{r gram-r}
# Pure-R gram_matrix (reproduces the original implementation)
gram_matrix_r <- function(s, h, alphas) {
  d <- nrow(s$points)
  Dm <- nrow(alphas)
  N <- ncol(s$points)
  U_sc <- s$points / h
  Phi <- matrix(NA_real_, Dm, N)
  for (j in seq_len(Dm)) {
    powered <- U_sc^alphas[j, ]
    Phi[j, ] <- matrixStats::colProds(powered)
  }
  weight <- (2 * h)^d / (s$n_total * h^d)
  weight * tcrossprod(Phi)
}
```

---

## Numerical parity

Before timing, we verify that both implementations agree to machine precision.

```{r parity}
set.seed(1L)
m <- 2L
alphas <- build_alphas(m, d)
N <- 1000L
pts <- matrix(runif(d * N, -h, h), d, N)
s <- list(points = pts, n_total = N)

B_r <- gram_matrix_r(s, h, alphas)
B_cpp <- gram_matrix_cpp(pts, h, alphas, N)

max(abs(B_r - B_cpp))
```

---

## Benchmark: varying N (m = 1, d = 2)

```{r bench-N, cache=TRUE}
set.seed(2L)
m <- 1L
alphas <- build_alphas(m, d)

results_N <- lapply(Ns, function(N) {
  pts <- matrix(runif(d * N, -h, h), d, N)
  s <- list(points = pts, n_total = N)
  bm <- bench::mark(
    R = gram_matrix_r(s, h, alphas),
    Cpp = gram_matrix_cpp(pts, h, alphas, N),
    check = FALSE, iterations = 50L
  )
  data.frame(
    N = N, expr = c("R", "Cpp"),
    median_us = as.numeric(bm$median) * 1e6
  )
})
bench_N <- do.call(rbind, results_N)
bench_N$expr <- as.character(bench_N$expr)
```

```{r plot-N, fig.cap="Median time (µs) vs N, m = 1, d = 2"}
pal <- c(R = "#EF4444", Cpp = "#3B82F6")
bench_R <- bench_N[bench_N$expr == "R", ]
bench_Cpp <- bench_N[bench_N$expr == "Cpp", ]

plot(bench_R$N, bench_R$median_us,
  type = "b", pch = 16, lwd = 2, col = pal["R"],
  log = "xy", xlab = "N (quadrature points)", ylab = "Median time (µs)",
  main = "gram_matrix: R vs C++  (m = 1, d = 2)",
  xaxt = "n",
  ylim = range(c(bench_R$median_us, bench_Cpp$median_us))
)
axis(1, at = Ns, labels = format(Ns, big.mark = ","))
lines(bench_Cpp$N, bench_Cpp$median_us,
  type = "b", pch = 16, lwd = 2, col = pal["Cpp"]
)
legend("topleft",
  legend = c("R", "C++"), col = pal, pch = 16, lty = 1,
  lwd = 2, bty = "n"
)
```

```{r speedup-N}
speedup <- bench_R$median_us / bench_Cpp$median_us
data.frame(N = Ns, speedup = round(speedup, 1L))
```

---

## Benchmark: varying m (N = 2000, d = 2)

```{r bench-m, cache=TRUE}
set.seed(3L)
N <- 2000L
pts <- matrix(runif(d * N, -h, h), d, N)
s <- list(points = pts, n_total = N)

results_m <- lapply(ms, function(m) {
  alphas <- build_alphas(m, d)
  bm <- bench::mark(
    R = gram_matrix_r(s, h, alphas),
    Cpp = gram_matrix_cpp(pts, h, alphas, N),
    check = FALSE, iterations = 50L
  )
  data.frame(
    m = m, Dm = nrow(alphas),
    expr = c("R", "Cpp"),
    median_us = as.numeric(bm$median) * 1e6
  )
})
bench_m <- do.call(rbind, results_m)
```

```{r plot-m, fig.cap="Median time (µs) vs degree m, N = 2000, d = 2"}
bench_mR <- bench_m[bench_m$expr == "R", ]
bench_mCpp <- bench_m[bench_m$expr == "Cpp", ]
Dm_vals <- bench_mR$Dm

plot(Dm_vals, bench_mR$median_us,
  type = "b", pch = 16, lwd = 2, col = pal["R"],
  xlab = expression(D[m] ~ "(basis size)"), ylab = "Median time (µs)",
  main = "gram_matrix: R vs C++  (N = 2000, d = 2)",
  ylim = range(c(bench_mR$median_us, bench_mCpp$median_us))
)
lines(Dm_vals, bench_mCpp$median_us,
  type = "b", pch = 16, lwd = 2, col = pal["Cpp"]
)
legend("topleft",
  legend = c("R", "C++"), col = pal, pch = 16, lty = 1,
  lwd = 2, bty = "n"
)
axis(3, at = Dm_vals, labels = paste0("m=", ms), tick = FALSE)
```

```{r speedup-m}
data.frame(
  m = ms, Dm = bench_mR$Dm,
  speedup = round(bench_mR$median_us / bench_mCpp$median_us, 1L)
)
```

---

## End-to-end: full pipeline (`density_lp_point`), varying n

We now benchmark the **complete** point estimator: Gram matrix + Cholesky +
observation filtering + $\Phi$ evaluation + forward solves. The quadrature
samples ($N = 1\,000$) are pre-generated and fixed; only the observations $X$
vary.

Domain: unit square $[0,1]^2$, target $t = (0.5, 0.5)$ (pure interior, so
$V(h) = [0.2, 0.8]^2 \subset [0,1]^2$ and all quadrature points are accepted).

```{r e2e-def}
build_Phi <- densityLP:::build_Phi
lp_estimator_cpp <- densityLP:::lp_estimator_cpp

density_lp_point_r <- function(X, t, h, s, alphas) {
  d <- ncol(X)
  n <- nrow(X)
  B <- gram_matrix_r(s, h, alphas)
  L <- chol(B)
  e1 <- numeric(nrow(alphas))
  e1[1L] <- 1.0
  H_0 <- forwardsolve(t(L), e1)
  U <- sweep(X, 2L, t, "-")
  in_V <- apply(abs(U) <= h, 1L, all)
  U_V <- t(U[in_V, , drop = FALSE])
  if (ncol(U_V) == 0L) {
    return(0)
  }
  Phi_V <- build_Phi(U_V, h, alphas)
  H_V <- forwardsolve(t(L), Phi_V)
  (h^(-d) / n) * sum(H_0 * H_V)
}

density_lp_point_cpp <- function(X, t, h, s, alphas) {
  U_obs <- t(sweep(X, 2L, t, "-"))
  lp_estimator_cpp(s$points, s$n_total, U_obs, h, alphas)
}

is_square <- function(x, y) x >= 0 & x <= 1 & y >= 0 & y <= 1
dom_sq <- domain_from_indicator(is_square)
t_ctr <- c(0.5, 0.5)
N_q <- 1000L
set.seed(10L)
sampler_sq <- dom_sq$sampler_factory()
s_ctr <- sampler_sq(N_q, t_ctr, h)
```

```{r e2e-bench-n, cache=TRUE}
set.seed(11L)
ns_e2e <- c(200L, 500L, 1000L, 2000L, 5000L, 10000L)
alphas_e2e <- build_alphas(1L, d)

res_n_e2e <- lapply(ns_e2e, function(n) {
  X <- matrix(runif(d * n), n, d)
  bm <- bench::mark(
    R = density_lp_point_r(X, t_ctr, h, s_ctr, alphas_e2e),
    Cpp = density_lp_point_cpp(X, t_ctr, h, s_ctr, alphas_e2e),
    check = FALSE, iterations = 50L
  )
  data.frame(
    n = n, expr = c("R", "Cpp"),
    median_us = as.numeric(bm$median) * 1e6
  )
})
bench_n_e2e <- do.call(rbind, res_n_e2e)
```

```{r e2e-plot-n, fig.cap="Median time (µs) vs n, full pipeline, m = 1, d = 2"}
bench_R_e2e <- bench_n_e2e[bench_n_e2e$expr == "R", ]
bench_Cpp_e2e <- bench_n_e2e[bench_n_e2e$expr == "Cpp", ]

plot(bench_R_e2e$n, bench_R_e2e$median_us,
  type = "b", pch = 16, lwd = 2, col = pal["R"],
  log = "xy", xlab = "n (observations)", ylab = "Median time (µs)",
  main = "density_lp_point: R vs C++  (m = 1, d = 2)",
  xaxt = "n",
  ylim = range(c(bench_R_e2e$median_us, bench_Cpp_e2e$median_us))
)
axis(1, at = ns_e2e, labels = format(ns_e2e, big.mark = ","))
lines(bench_Cpp_e2e$n, bench_Cpp_e2e$median_us,
  type = "b", pch = 16, lwd = 2, col = pal["Cpp"]
)
legend("topleft",
  legend = c("R", "C++"), col = pal, pch = 16, lty = 1,
  lwd = 2, bty = "n"
)
```

```{r e2e-speedup-n}
data.frame(
  n = ns_e2e,
  speedup = round(bench_R_e2e$median_us / bench_Cpp_e2e$median_us, 1L)
)
```

---

## End-to-end: full pipeline, varying m (n = 2 000, d = 2)

```{r e2e-bench-m, cache=TRUE}
set.seed(12L)
n_e2e <- 2000L
X_e2e <- matrix(runif(d * n_e2e), n_e2e, d)

res_m_e2e <- lapply(ms, function(m) {
  alphas <- build_alphas(m, d)
  set.seed(13L)
  s <- sampler_sq(N_q, t_ctr, h)
  bm <- bench::mark(
    R = density_lp_point_r(X_e2e, t_ctr, h, s, alphas),
    Cpp = density_lp_point_cpp(X_e2e, t_ctr, h, s, alphas),
    check = FALSE, iterations = 50L
  )
  data.frame(
    m = m, Dm = nrow(alphas), expr = c("R", "Cpp"),
    median_us = as.numeric(bm$median) * 1e6
  )
})
bench_m_e2e <- do.call(rbind, res_m_e2e)
```

```{r e2e-plot-m, fig.cap="Median time (µs) vs degree m, full pipeline, n = 2 000, d = 2"}
bench_mR_e2e <- bench_m_e2e[bench_m_e2e$expr == "R", ]
bench_mCpp_e2e <- bench_m_e2e[bench_m_e2e$expr == "Cpp", ]
Dm_e2e <- bench_mR_e2e$Dm

plot(Dm_e2e, bench_mR_e2e$median_us,
  type = "b", pch = 16, lwd = 2, col = pal["R"],
  xlab = expression(D[m] ~ "(basis size)"), ylab = "Median time (µs)",
  main = "density_lp_point: R vs C++  (n = 2 000, d = 2)",
  ylim = range(c(bench_mR_e2e$median_us, bench_mCpp_e2e$median_us))
)
lines(Dm_e2e, bench_mCpp_e2e$median_us,
  type = "b", pch = 16, lwd = 2, col = pal["Cpp"]
)
legend("topleft",
  legend = c("R", "C++"), col = pal, pch = 16, lty = 1,
  lwd = 2, bty = "n"
)
axis(3, at = Dm_e2e, labels = paste0("m=", ms), tick = FALSE)
```

```{r e2e-speedup-m}
data.frame(
  m = ms, Dm = bench_mR_e2e$Dm,
  speedup = round(bench_mR_e2e$median_us / bench_mCpp_e2e$median_us, 1L)
)
```
