---
title: "Convergence study on complex domains"
format:
  html:
    toc: true
    fig-width: 9
    fig-height: 6
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Convergence study on complex domains}
  %\VignetteEngine{quarto::quarto}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(densityLP)
```

We illustrate how the local polynomial estimator ($m = 1$, bandwidth $h = 0.3$)
behaves at points with varying proximity to the boundary, on two domains:

1. the unit cube $[0, 1]^3$ in $\mathbb{R}^3$,
2. the polynomial sector
   $D_2 = \{(x, y) : x \in [0, 1],\; 0 \le y \le x^2\}$ in $\mathbb{R}^2$.

For each domain we simulate $R = 100$ independent datasets of sizes
$n \in \{100, 500, 1000, 5000\}$ from the uniform distribution on the domain,
compute $\hat f(t)$ at several target points $t$, and summarise the
distribution of $\hat f(t)$ as well as the mean squared error
$\mathrm{MSE}(n, t) = \mathbb{E}[(\hat f(t) - f(t))^2]$.

---

## Common parameters

```{r params}
ns     <- c(100L, 500L, 1000L, 5000L)
R      <- 100L   # Monte Carlo replications
m      <- 1L     # local linear
h      <- 0.3    # bandwidth (shared across both domains)
N_quad_sector <- 500L   # quadrature points per target
N_quad_cube <- 5000L   # quadrature points per target
```

## Plotting helpers

```{r helpers}
# Colour palette (one per evaluation point)
pal <- c("#3B82F6", "#F59E0B", "#EF4444", "#22C55E")

# Bias (mean - f_true) +/- 1 sd across replications, one panel per target point
plot_convergence <- function(res, ns, f_true, pt_labels, title) {
  n_pts <- length(pt_labels)
  nc    <- min(n_pts, 2L)
  nr    <- ceiling(n_pts / nc)
  par(mfrow = c(nr, nc), mar = c(4, 4.5, 3, 1), oma = c(0, 0, 2, 0))

  for (k in seq_len(n_pts)) {
    vals  <- res[, , k]                         # R x length(ns)
    bias  <- colMeans(vals, na.rm = TRUE) - f_true
    sigma <- apply(vals, 2L, sd, na.rm = TRUE)
    ylim  <- range(c(bias - sigma, bias + sigma, 0), na.rm = TRUE)

    plot(
      ns, bias,
      type = "b", pch = 16, lwd = 2, col = pal[k],
      log  = "x", xlab = "n",
      ylab = expression(hat(f)(t) - f(t)),
      main = pt_labels[k], ylim = ylim, xaxt = "n"
    )
    axis(1, at = ns, labels = format(ns, big.mark = ","))
    polygon(
      c(ns, rev(ns)),
      c(bias + sigma, rev(bias - sigma)),
      col = adjustcolor(pal[k], alpha.f = 0.25), border = NA
    )
    abline(h = 0, lty = 2, lwd = 1.5, col = "#64748B")
  }
  mtext(title, outer = TRUE, cex = 1.1, font = 2)
  par(mfrow = c(1L, 1L))
}

# MSE curves on log-log scale
plot_mse <- function(res, ns, f_true, pt_labels, title) {
  mse <- apply(res, c(2L, 3L), function(x) mean((x - f_true)^2, na.rm = TRUE))
  # mse: length(ns) x n_pts
  matplot(
    ns, mse,
    type = "b", pch = 16, lwd = 2,
    col  = pal[seq_len(length(pt_labels))],
    log  = "xy", xlab = "n", ylab = "MSE",
    main = title, xaxt = "n"
  )
  axis(1, at = ns, labels = format(ns, big.mark = ","))
  legend(
    "topright",
    legend = pt_labels,
    col    = pal[seq_len(length(pt_labels))],
    pch    = 16, lty = 1, lwd = 2, bty = "n", cex = 0.9
  )
}
```

---

## 1. Cube $[0, 1]^3$

### Domain and target points

The underlying density is uniform: $f \equiv 1$ on $[0, 1]^3$.
We evaluate $\hat f$ at four points that progressively approach the boundary:

| Label | $t$ | Position |
|---|---|---|
| Center | $(0.5, 0.5, 0.5)$ | interior |
| Face center | $(0, 0.5, 0.5)$ | face midpoint |
| Edge midpoint | $(0, 0, 0.5)$ | edge midpoint |
| Corner | $(0, 0, 0)$ | vertex |

```{r cube-setup}
is_cube  <- function(x, y, z) x >= 0 & x <= 1 & y >= 0 & y <= 1 & z >= 0 & z <= 1
dom_cube <- domain_from_indicator(is_cube)
f_cube   <- 1   # true density

t_cube <- rbind(
  center = c(0.5, 0.5, 0.5),
  face   = c(0.0, 0.5, 0.5),
  edge   = c(0.0, 0.0, 0.5),
  corner = c(0.0, 0.0, 0.0)
)
pt_cube <- c("Center", "Face center", "Edge midpoint", "Corner")
```

### Monte Carlo simulation

```{r cube-sim, cache=TRUE}
set.seed(2024L)
res_cube <- array(
  NA_real_,
  dim      = c(R, length(ns), nrow(t_cube)),
  dimnames = list(NULL, as.character(ns), pt_cube)
)

for (r in seq_len(R)) {
  # Nested samples: use first n rows of a single large dataset
  X_full <- matrix(runif(max(ns) * 3L), max(ns), 3L)
  for (j in seq_along(ns)) {
    X   <- X_full[seq_len(ns[j]), , drop = FALSE]
    fit <- density_lp(X, t_grid = t_cube, h = h, m = m,
                      domain = dom_cube, N_quad = N_quad_cube)
    res_cube[r, j, ] <- fit$estimate
  }
}
```

### Results

```{r cube-conv-plot, fig.height=8, fig.cap="Cube: bias and ± 1σ by sample size"}
plot_convergence(res_cube, ns, f_cube, pt_cube,
                 "Cube [0,1]³ — bias ± 1σ")
```

The central line shows the empirical bias $\mathbb{E}[\hat f(t)] - f(t)$ and the
ribbon covers $\pm 1\sigma$. Both shrink as $n$ grows. Points near the boundary
(face, edge, corner) exhibit larger bias and variance because $V(h)$ is truncated.

```{r cube-mse-plot, fig.height=4, fig.cap="Cube: MSE on log-log scale"}
plot_mse(res_cube, ns, f_cube, pt_cube, "Cube [0,1]³ — MSE")
```

All four curves decrease roughly in parallel, but boundary points maintain a
higher MSE level for all $n$.

---

## 2. Polynomial sector $D_2$

### Domain and target points

$D_2 = \{(x, y) : x \in [0, 1],\; 0 \le y \le x^2\}$ has area $1/3$, so the
uniform density is $f \equiv 3$ on $D_2$.

| Label | $t$ | Position |
|---|---|---|
| Interior | $(0.6, 0.2)$ | well inside ($0.2 < 0.6^2 = 0.36$) |
| Boundary ($y = 0$) | $(0.5, 0)$ | lower edge |
| Apex | $(0, 0)$ | singular corner |

```{r sector-setup}
dom_sector2 <- domain_sector(2L)
f_sector    <- 3   # true density (= 1 / area(D_2) = 1 / (1/3))

t_sector <- rbind(
  interior = c(0.6, 0.2),
  boundary = c(0.5, 0.0),
  apex     = c(0.0, 0.0)
)
pt_sector <- c("Interior", "Boundary (y = 0)", "Apex (0, 0)")
```

### Monte Carlo simulation

```{r sector-sim-fun}
# Uniform draws on D_2 via rejection from [0,1]^2 (acceptance rate = 1/3)
sim_sector <- function(n) {
  acc <- matrix(NA_real_, 0L, 2L)
  while (nrow(acc) < n) {
    need <- ceiling(3.5 * (n - nrow(acc)))
    U    <- matrix(runif(need * 2L), need, 2L)
    ok   <- U[, 2L] <= U[, 1L]^2
    acc  <- rbind(acc, U[ok, , drop = FALSE])
  }
  acc[seq_len(n), ]
}
```

```{r sector-sim, cache=TRUE}
set.seed(2025L)
res_sector <- array(
  NA_real_,
  dim      = c(R, length(ns), nrow(t_sector)),
  dimnames = list(NULL, as.character(ns), pt_sector)
)

for (r in seq_len(R)) {
  X_full <- sim_sector(max(ns))
  for (j in seq_along(ns)) {
    X   <- X_full[seq_len(ns[j]), , drop = FALSE]
    fit <- density_lp(X, t_grid = t_sector, h = h, m = m,
                      domain = dom_sector2, N_quad = N_quad_sector)
    res_sector[r, j, ] <- fit$estimate
  }
}
```

### Results

```{r sector-conv-plot, fig.height=5, fig.cap="Sector: bias and ± 1σ by sample size"}
plot_convergence(res_sector, ns, f_sector, pt_sector,
                 "Sector D₂ — bias ± 1σ")
```

At the apex $(0, 0)$, the domain degenerates ($V(h)$ has volume $h^3/3 \approx 0.009$),
so very few data points contribute for small $n$, leading to high bias and variance.
The estimator remains consistent but convergence is slow.

```{r sector-mse-plot, fig.height=4, fig.cap="Sector: MSE on log-log scale"}
plot_mse(res_sector, ns, f_sector, pt_sector, "Sector D₂ — MSE")
```

The interior point converges fastest. The apex exhibits the slowest convergence,
illustrating the theoretical result that boundary regions with vanishing volume
require larger $n$ to achieve the same precision.
