---
title: "Bandwidth and degree selection by LOO cross-validation"
format:
  html:
    toc: true
    fig-width: 8
    fig-height: 5
    code-fold: true
vignette: >
  %\VignetteIndexEntry{Bandwidth and degree selection by LOO cross-validation}
  %\VignetteEngine{quarto::quarto}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(densityLP)
```

`cv_density_lp()` selects the bandwidth $h$ and polynomial degree $m$ jointly
by minimising the leave-one-out log-likelihood score

$$\mathrm{CV}(m, h) = -\frac{1}{n}\sum_{i=1}^n \log \hat f_{\gamma,-i}(X_i),$$

where the LOO estimate uses the algebraic correction
$$\hat f_{\gamma,-i}(X_i)
  = \frac{n\,\hat f_\gamma(X_i) - h^{-d}\|H_0^{(X_i)}\|^2}{n-1}$$
and $\|H_0^{(X_i)}\|^2 = (B_{X_i}^{-1})_{11}$ is the self-influence scalar
returned as a byproduct of the C++ pipeline. The full score matrix
`cv$scores` ($|\mathcal{M}| \times |\mathcal{H}|$) is stored, enabling
diagnostic plots of the CV surface.

---

## 1. Uniform density on the unit square

The true density is $f \equiv 1$ on $[0,1]^2$. A flat density is best
estimated with a large bandwidth, so CV should select the largest $h$ in the
grid and prefer $m = 0$ (local constant).

```{r square-setup}
set.seed(1L)
n      <- 2000L
X_sq   <- matrix(runif(n * 2L), n, 2L)

is_sq  <- function(M) apply(M, 1L, function(x) all(x >= 0 & x <= 1))
dom_sq <- domain_from_indicator(is_sq, d = 2L)

h_grid <- seq(0.02, 0.5, by = 0.02)
m_grid <- 0:2
```

```{r square-cv, cache=TRUE}
cv_sq <- cv_density_lp(X_sq, h_grid = h_grid, m_grid = m_grid,
                       domain = dom_sq, N_quad = 300L)
cv_sq
```

```{r square-cv-plot, fig.cap="CV score vs bandwidth for each degree (uniform density). The star marks the selected pair."}
pal <- c("m=0" = "#EF4444", "m=1" = "#3B82F6", "m=2" = "#22C55E")

scores_plot <- cv_sq$scores
scores_plot[!is.finite(scores_plot)] <- NA

ylim <- range(scores_plot, na.rm = TRUE)

plot(cv_sq$h_grid, scores_plot["m=0", ],
     type = "b", pch = 16, lwd = 2, col = pal["m=0"],
     xlab = "h", ylab = "CV score  (lower = better)",
     main = "LOO CV — unit square, f ≡ 1",
     ylim = ylim)
lines(cv_sq$h_grid, scores_plot["m=1", ],
      type = "b", pch = 16, lwd = 2, col = pal["m=1"])
lines(cv_sq$h_grid, scores_plot["m=2", ],
      type = "b", pch = 16, lwd = 2, col = pal["m=2"])

abline(v = cv_sq$h_hat, lty = 2, col = "grey50")
best <- cv_sq$scores[paste0("m=", cv_sq$m_hat),
                     as.character(round(cv_sq$h_hat, 6L))]
points(cv_sq$h_hat, best, pch = 8, cex = 2, lwd = 2, col = "black")
legend("topright", legend = paste0("m = ", m_grid),
       col = pal, pch = 16, lty = 1, lwd = 2, bty = "n")
```

CV correctly identifies that a flat density benefits from a large bandwidth.
The $m = 2$ curve has gaps where the LOO denominator was negative (V(h) too
small for the quadrature budget at that observation).

```{r square-fit, cache=TRUE}
t_grid_sq <- as.matrix(expand.grid(
  x = seq(0.05, 0.95, length.out = 25L),
  y = seq(0.05, 0.95, length.out = 25L)
))

fit_sq <- density_lp(X_sq, t_grid = t_grid_sq,
                     h = cv_sq$h_hat, m = cv_sq$m_hat,
                     domain = dom_sq, N_quad = 500L)
```

```{r square-fit-plot, fig.cap="Estimated density with CV-selected parameters. True value: f ≡ 1."}
plot(fit_sq,
     main = sprintf("Estimate  (m = %d, h = %.2f) — true f ≡ 1",
                    cv_sq$m_hat, cv_sq$h_hat))
```

---

## 2. Non-uniform density on the unit square

We now simulate $n = 200$ observations from
$f(x, y) = f_{\mathrm{B}(2,5)}(x)\,f_{\mathrm{B}(2,5)}(y)$,
a density concentrated near $(0.3, 0.3)$ with a sharp drop towards $(1,1)$.
A smaller bandwidth is needed to track the gradient.

```{r beta-setup}
set.seed(2L)
X_bt <- cbind(rbeta(n, 2, 5), rbeta(n, 2, 5))
```

```{r beta-cv, cache=TRUE}
cv_bt <- cv_density_lp(X_bt, h_grid = h_grid, m_grid = m_grid,
                       domain = dom_sq, N_quad = 300L)
cv_bt
```

```{r beta-cv-plot, fig.cap="CV score vs bandwidth (Beta(2,5) marginals). The selected h is smaller than for the uniform case."}
scores_bt <- cv_bt$scores
scores_bt[!is.finite(scores_bt)] <- NA

ylim_bt <- range(scores_bt, na.rm = TRUE)

plot(cv_bt$h_grid, scores_bt["m=0", ],
     type = "b", pch = 16, lwd = 2, col = pal["m=0"],
     xlab = "h", ylab = "CV score  (lower = better)",
     main = "LOO CV — unit square, f = Beta(2,5) × Beta(2,5)",
     ylim = ylim_bt)
lines(cv_bt$h_grid, scores_bt["m=1", ],
      type = "b", pch = 16, lwd = 2, col = pal["m=1"])
lines(cv_bt$h_grid, scores_bt["m=2", ],
      type = "b", pch = 16, lwd = 2, col = pal["m=2"])

abline(v = cv_bt$h_hat, lty = 2, col = "grey50")
best_bt <- cv_bt$scores[paste0("m=", cv_bt$m_hat),
                        as.character(round(cv_bt$h_hat, 6L))]
points(cv_bt$h_hat, best_bt, pch = 8, cex = 2, lwd = 2, col = "black")
legend("topright", legend = paste0("m = ", m_grid),
       col = pal, pch = 16, lty = 1, lwd = 2, bty = "n")
```

```{r beta-fit, cache=TRUE}
fit_bt <- density_lp(X_bt, t_grid = t_grid_sq,
                     h = cv_bt$h_hat, m = cv_bt$m_hat,
                     domain = dom_sq, N_quad = 500L)
```

```{r beta-compare, fig.width=10, fig.height=4, fig.cap="Left: true density. Right: CV-selected estimate."}
f_true_bt <- function(x, y) dbeta(x, 2, 5) * dbeta(y, 2, 5)

par(mfrow = c(1L, 2L))

xs <- seq(0.05, 0.95, length.out = 25L)
ys <- xs
Z_true <- outer(xs, ys, f_true_bt)
image(xs, ys, Z_true,
      xlab = "x", ylab = "y",
      main = "True density  f = B(2,5) × B(2,5)")

plot(fit_bt,
     main = sprintf("Estimate  (m = %d, h = %.2f)", cv_bt$m_hat, cv_bt$h_hat))

par(mfrow = c(1L, 1L))
```

The CV correctly shifts to a smaller bandwidth for the structured density,
and the estimate tracks the concentration near $(0.3, 0.3)$.
